Characterizing the response of subarctic European vegetation to key environmental variables for land surface modeling of ozone pollution and risk assessment

### 2021-10-15 VegOzone.tex
Checked

\begin{table}[t]
  \caption{Start and end of growing season in \unit{doy}.}
  \label{tab:sensitivity_tests_gs}
  \begin{tabular}{lcll}
    \tophline
    Species & Year & $A_\mathrm{start}$ & $A_\mathrm{end}$\\
    \middlehline
    Default$^b$ & -- & 100 & 307\\
    \multirow{2}{*}{Deciduous tree} & 2018 & 129$^*$ & 261$^a$ \\
    & 2019 & 130$^c$ & 274$^a$ \\
    \multirow{2}{*}{Coniferous tree} & 2018 & 122$^a$ & 261$^a$ \\
    & 2019 & 106$^a$ & 274$^a$ \\
    \multirow{2}{*}{Perennial grassland} & 2018 & 159$^d$ & 261$^a$\\
    & 2019 & 161$^d$ & 274$^a$ \\
    \bottomhline
  \end{tabular}
  \belowtable{$^a$ MODIS (Aqua/Terra) net photosynthesis product;\\
    $^b$ \citet{ACP:Simpson2012},\\
    $^c$ $5\,\unit{days}-5\,\unit{^\circ C}$-rule, $T_\mathrm{air}$ from \href{seNorge.no}{seNorge.no};\\
    $^d$ One month after snow melt; reference station Øvre Neiden (Sør-Varanger, NOR)} % Table Footnotes
\end{table}

@@

Based on a MODIS photosynthesis product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The determined start of growing season in 2018 ($\unit{doy}~122$) and 2019 ($\unit{doy}~106$) lies within the range given by observations in Fennoscandia \citep{TB:Kolari2007,IVL:Karlsson2018}. We deduced the start of the growing season for deciduous trees based on gridded observational temperature and for perennial grassland on snow depth data. For perennial grassland, we assumed a one month latency period after snow melt that is supported by observational data from Rovaniemi (Finland) \citep{FCR:Korhonen2018}. With respect to ongoing climate change, a clear positive trend emerged in length ($5.2\,\unit{d\,decade^{-1}}$) of the growing season that is almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) (Appendix Fig.~\ref{fig:greening_season_change_Svanvik}).

 The $\mathrm{DO_3SE}$ modeling results are in contrast to our observation of more visible damage in 2018 compared to 2019. However, visible damage caused by peak ozone uptake and a dependence of ozone sensibility on phenology are not accounted for in $\mathrm{POD_y}$. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible damage explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. In particular, forbs and perennial grassland have been found to be more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed damage in the ozone garden to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

At the same time, intra-species variability even at the same location is typically non-negligible \citep{EP:Bassin2004} but not accounted for in the models. The literature results regarding ozone sensitivity of natural vegetation in northern Europe are contratictive. \citet{FS:Subramanian2014} reported a higher, modeled reduction of net primary production in conifers ($4.3-15.5\,\unit{\%}$) than birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, while \citet{Amb:Girgzdiene2009} observed more visible damage on deciduous trees than on Norway spruce in Lithuania. But visible damage and biomass reduction may not occur at the same rate and hence may not be interchangeable.

In our study, we do not explicitly account for this natural distribution of individual plant traits, but our bespoke response functions reflect the diversity within a given species. In particular, we assumed two tolerance levels to cold temperatures than for central European species. To this end, we computed a PDF from observed temperature at Svanhovd (May--September, 1992--2000). We adjusted the parameters in the temperature response function of stomatal conductance to increase the enclosed area with the PDF. Assuming a proportionality between $g_\mathrm{sto}$ and $A_\mathrm{n}$, this is equivalent to optimizing carbon uptake. Similarly, we deduced a PPFD PDF and varied the light response function to test a higher or lower extent of stomatal opening at low light intensities. Remark that these parameterizations are hypothetical and need verification by experimental data. We found that soil water potential under 2018/19 meteorological conditions was negligible when considering the bespoke growing season. We have not assessed the effect of the plants' response to VPD in detail, but our climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime.

@@

In this regard, vegetation might become more exposed to higher ozone concentrations in early phenological stages due to an overlap with ozone spring peak \citep{EP:Karlsson2007}. However, the decline of this ozone spring peak is partly caused by uptake of vegetation. It, therefore, remains unclear whether an earlier start of the growing season will increase the exposure of vegetation to ozone or lead to a compensation due to an earlier decline of the peak ozone in the future. \citet{ESPR:Hayes2021} pointed out that the highest susceptibility of vegetation to future ozone variability is likely to occur in summer when vegetation is most productive. The actual resilience of northern Fennoscandia plant communities to climate change strongly depend on both the range of existing acclimations as well as the actual type acclimation. This means vegetation that is acclimated to cold temperatures and little to heat- and drought stress might suffer more strongly from a more frequent occurance of heatwaves than plants which have a higher temperature tolerance.

Following the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and found substantial reductions for deciduous trees ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) in both years. These reductions are higher than for the default parameterizations, $11.2\,\unit{\%}$ (deciduous) and $9.6\,\unit{\%}$ (perennial grassland). For coniferous trees, the estimated reduction in biomass differs insignificantly between our bespoke parameterizations ($2.5-3\,\unit{\%}$) and the default ($2.31\,\unit{\%}$). This indicates a higher modeled robustness to ozone damage in coniferous trees. All these results have to be treated with care because the biomass reduction functions were established based on studies from less extreme climates. In particular, CLs for deciduous and coniferous trees were breached in all years and for all parameterizations. For perennial grassland, the CL was not breached if a bespoke GS was assumed, but the mapping manual defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constrains on productivity leading to more severe economical consequences.

Beyond the risk assessment of ozone-induced damage, our bespoke temperature response functions, once verified by observations, can have important implications on land-surface modeling in global models, where problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. We pointed out, with respect to default perennial grassland parameterization, that these perennial grasslands did not show any substantial stomatal conductance in 2019 characterized as a normal year. In terms of GPP this would mean unrealistically low carbon uptake. Automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.

>>>>

\begin{table}[t]
  \caption{Start and end of growing season in \unit{doy}.}
  \label{tab:sensitivity_tests_gs}
  \begin{tabular}{lcll}
    \tophline
    Species & Year & $A_\mathrm{start}$ & $A_\mathrm{end}$\\
    \middlehline
    Default$^b$ & -- & 100 & 307\\
    \multirow{2}{*}{Deciduous tree} & 2018 & 129$^*$ & 261$^a$ \\
    & 2019 & 130$^c$ & 274$^a$ \\
    \multirow{2}{*}{Coniferous tree} & 2018 & 122$^a$ & 261$^a$ \\
    & 2019 & 106$^a$ & 274$^a$ \\
    \multirow{2}{*}{Perennial grassland} & 2018 & 159$^d$ & 261$^a$\\
    & 2019 & 161$^d$ & 274$^a$ \\
    \bottomhline
  \end{tabular}
  \belowtable{$^a$ MODIS (Aqua/Terra) net photosynthesis product;\\
    $^b$ \citet{ACP:Simpson2012},\\
    $^c$ $5\,\unit{days}-5\,\unit{^\circ C}$-rule, $T_\mathrm{air}$ from \href{seNorge.no}{seNorge.no};\\
    $^d$ One month after snowmelt; reference station Øvre Neiden (Sør-Varanger, NOR)} % Table Footnotes
\end{table}

@@

Based on a MODIS photosynthesis product \citep{MODIS_PSN}, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The determined start of growing season in 2018 ($\unit{doy}~122$) and 2019 ($\unit{doy}~106$) lies within the range given by observations in Fennoscandia \citep{TB:Kolari2007,IVL:Karlsson2018}. We deduced the start of the growing season for deciduous trees based on gridded observational temperature and for perennial grassland on snow depth data. For perennial grassland, we assumed an one-month latency period after snowmelt that is supported by observational data from Rovaniemi (Finland) \citep{FCR:Korhonen2018}. With respect to ongoing climate change, a clear positive trend emerged in length ($5.2\,\unit{d\,decade^{-1}}$) of the growing season that is almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) (Appendix Fig.~\ref{fig:greening_season_change_Svanvik}).

 The $\mathrm{DO_3SE}$ modeling results are in contrast to our observation of more visible damage in 2018 compared to 2019. However, visible damage caused by peak ozone uptake and dependence of ozone sensibility on phenology are not accounted for in $\mathrm{POD_y}$. In this regard, open-top chamber (OTC) experiments performed in northern Finland where Scots pine and downy birch seedlings were exposed to elevated ozone concentrations attributed a reduction in biomass and reproduction with visible damage explicitly to peak \chem{O_3} concentrations and fast phenological development at high growth rate \citep{Amb:Manninen2009}. In particular, forbs and perennial grassland are more susceptible to ozone-induced damage in the reproductive state \citep{EP:Bassin2004}. 
That might explain why we observed damage in the ozone garden to a larger extent in 2018 than in 2019. An associated damage function to ozone depending on leaf age as proposed by \citet{AE:Musselman2006} or phenological state might improve the predictions.

At the same time, intra-species variability even at the same location is typically non-negligible \citep{EP:Bassin2004} but not accounted for in the models. The literature results regarding the ozone sensitivity of natural vegetation in northern Europe are contradictive. \citet{FS:Subramanian2014} reported a higher, modeled reduction of net primary production in conifers ($4.3-15.5\,\unit{\%}$) than birch ($1.4-4.3\,\unit{\%}$) under elevated ozone in Sweden, while \citet{Amb:Girgzdiene2009} observed more visible damage on deciduous trees than on Norway spruce in Lithuania. But visible damage and biomass reduction may not occur at the same rate and hence may not be interchangeable.

In our study, we do not explicitly account for this natural distribution of individual plant traits, but our bespoke response functions reflect the diversity within a given species. In particular, we assumed two tolerance levels to cold temperatures than for central European species. To this end, we computed a PDF from observed temperature at Svanhovd (May--September 1992--2000). We adjusted the parameters in the temperature response function of stomatal conductance to increase the enclosed area with the PDF. Assuming a proportionality between $g_\mathrm{sto}$ and $A_\mathrm{n}$, this is equivalent to optimizing carbon uptake. Similarly, we deduced a PPFD PDF and varied the light response function to test a higher or lower extent of stomatal opening at low light intensities. Remark that these parameterizations are hypothetical and need verification by experimental data. We found that soil water potential under 2018/19 meteorological conditions was negligible when considering the bespoke growing season. We have not assessed the effect of the plants' response to VPD in detail, but our climatological data indicated that conditions at Svanhovd are shifting towards a more VPD limited regime.

@@

In this regard, vegetation might become more exposed to higher ozone concentrations in early phenological stages due to an overlap with ozone spring peak \citep{EP:Karlsson2007}. However, the decline of this ozone spring peak is partly caused by the uptake of vegetation. It, therefore, remains unclear whether an earlier start of the growing season will increase the exposure of vegetation to ozone or lead to compensation due to an earlier decline of the peak ozone in the future. \citet{ESPR:Hayes2021} pointed out that the highest susceptibility of vegetation to future ozone variability is likely to occur in summer when vegetation is most productive. The actual resilience of northern Fennoscandia plant communities to climate change strongly depends on both the range of existing acclimations as well as the actual type of acclimation. This means vegetation that is acclimated to cold temperatures and little to heat- and drought stress might suffer more strongly from a more frequent occurence of heatwaves than plants that have a higher temperature tolerance.

Following the methods described in \citet{ICP:MappingManual2017}, we estimated biomass reductions due to ozone uptake for both 2018 and 2019 and found substantial reductions for deciduous trees ($15.5-17.4\,\unit{\%}$) and perennial grassland (above ground, $12.4-13.7\,\unit{\%}$) in both years. These reductions are higher than for the default parameterizations, $11.2\,\unit{\%}$ (deciduous) and $9.6\,\unit{\%}$ (perennial grassland). For coniferous trees, the estimated reduction in biomass differs insignificantly between our bespoke parameterizations ($2.5-3\,\unit{\%}$) and the default ($2.31\,\unit{\%}$). This indicates a higher modeled robustness to ozone damage in coniferous trees. All these results have to be treated with care because the biomass reduction functions were established based on studies from less extreme climates. In particular, CLs for deciduous and coniferous trees were breached in all years and for all parameterizations. For perennial grassland, the CL was not breached if a bespoke GS was assumed, but the mapping manual defined CL of $10\,\unit{\%}$ loss in biomass might not be transferable to subarctic regions with considerable constraints on productivity leading to more severe economical consequences.

Beyond the risk assessment of ozone-induced damage, our bespoke temperature response functions, once verified by observations, can have important implications on land-surface modeling in global models, where problems in productivity of species especially in the Arctic regions occur due to PFTs which are not suitable for the climate. We pointed out, concerning default perennial grassland parameterization, that these perennial grasslands did not show any substantial stomatal conductance in 2019 characterized as a normal year. In terms of GPP this would mean unrealistically low carbon uptake. Automation of the here proposed PDF-based acclimation using machine learning techniques could overcome these issues in the future.
